---
title: "Reproducible Research: Peer Assessment 1"
output: html_document
---
```{r}
library(knitr)
library(ggplot2)
library(dplyr)
library(plyr)
```
Download the file and unzip
Read the file

```{r}
temp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", temp, mode="wb")
unzip(temp, "activity.csv")
activity <- read.csv("activity.csv",header=T)
unlink(temp)
```
2.Process/transform the data (if necessary) into a format suitable for your analysis

What is mean total number of steps taken per day?

For this part of the assignment, you can ignore the missing values in the dataset.

```{r}
totalstepsperday <- aggregate(steps ~ date, data = activity, FUN = sum, na.rm = TRUE)
totalstepsperday
```
2. Make a histogram of the total number of steps taken each day
converting dates to Y-M-D format
```{r}
activity$date <- as.Date(activity$date, "%Y-%m-%d")
hist(totalstepsperday$steps, 
    main="Total Steps per Day", 
    xlab="Number of Steps per Day", 
    ylab = "Interval",
    col="seagreen",
    breaks=60)
```
3. Calculate and report the mean and median of the total number of steps taken per day
Mean
```{r}
msteps <- mean(totalstepsperday$steps)
msteps
```
Median 
```{r}
medsteps <- median(totalstepsperday$steps)
medsteps
```
check work using summary
```{r}
summary(totalstepsperday)
```

What is the average daily activity pattern?
1. Make a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
fivemin <- aggregate(steps ~ interval, data = activity, FUN = mean, na.rm = TRUE)
plot(x = fivemin$interval, 
    y = fivemin$steps, 
    type = "l", 
    col = "red",
    xlab = "5-minute Intervals",
    ylab = "Average Steps Taken ~ Days",
    main = "Average Daily Activity Pattern")
```
Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
maxsteps <- fivemin$interval[which.max(fivemin$steps)]
maxsteps
```
Imputing missing values

Note that there are a number of days/intervals where there are missing values (coded as \color{red}{\verb|NA|}NA). The presence of missing days may introduce bias into some calculations or summaries of the data

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with \color{red}{\verb|NA|}NAs)
```{r}
impute <- sum(is.na(activity$steps))
impute
```
2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
Replace NA values with the mean results for five minute intervals
3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
activity2 <- activity
nas <- is.na(activity2$steps)
avg_interval <- tapply(activity2$steps, activity2$interval, mean, na.rm=TRUE, simplify = TRUE)
activity2$steps[nas] <- avg_interval[as.character(activity2$interval[nas])]
names(activity2)
```
 Check for no-NA
```{r}
sum(is.na(activity2))
```
4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. 

Analysis without NAs:
```{r}
totalstepsperday2 <- aggregate(steps ~ date, data = activity2, FUN = sum, na.rm = TRUE)
totalstepsperday2
```
Histogram without NA values
```{r}
hist(totalstepsperday2$steps, 
    main = "Total Steps per Day (no-NA)", 
    xlab = "Number of Steps per Day", 
    ylab = "Interval",
    col="brown",
    breaks=60)
```

Do these values differ from the estimates from the first part of the assignment? 
What is the impact of imputing missing data on the estimates of the total daily number of steps?

Before Imputing the Data
```{r}
summary(totalstepsperday)
```
After Imputing the Data
```{r}
summary(totalstepsperday2)
```
Mean and median values are almost identical, but the quantiles are significantly different

Are there differences in activity patterns between weekdays and weekends?
For this part the \color{red}{\verb|weekdays()|}weekdays() function may be of some help here. Use the dataset with the filled-in missing values for this part.
```{r}
head(activity2)
```
1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.
```{r}
par(bg = 'blue')
activity2<- activity2%>%
        mutate(typeofday= ifelse(weekdays(activity2$date)=="Saturday" | weekdays(activity2$date)=="Sunday", "Weekend", "Weekday"))
head(activity2)
```
2. Make a panel plot containing a time series plot (i.e. \color{red}{\verb|type = "l"|}type="l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 
```{r}
fivemin2<- aggregate(steps ~ interval, data = activity2, FUN = mean, na.rm = TRUE)
head(fivemin2)

ggplot(activity2, aes(x =interval , y=steps, color=typeofday)) +
       geom_line() +
       labs(title = "Ave Daily Steps (type of day)", x = "Interval", y = "Total Number of Steps") +
       facet_wrap(~ typeofday, ncol = 1, nrow=2)
```





